---
title: "compareScenariosLimes"
author: "Christof Schötz"
output: 
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{compareScenariosLimes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval=FALSE
)
```

The function `compareScenariosLimes()` of the `limes` package is based on the `compareScenariosLimes()` function from the `remind2` [package](https://github.com/pik-piam/remind2/) and adapts it to the needs of the **LIMES** model. It can be used to visually compare the results of multiple runs of the **LIMES** model. See also `vignette("compareScenarios2")` from which this vignette is largely copied. 

It can be called via the `output.R` script in the **LIMES** model: Execute `Rscript output.R` in the **LIMES** folder, select `Comparison across runs` and then `compareScenariosLimes`.

The function reads the results from the output-mif-files (in the **LIMES** folder after runs are completed, i.e., the files `output/<scenario_folder>/LIMES_generic_<scenario>.mif`). Using this data, a document containing many plots is created and written as a PDF- or HTML-file. See `?limes::compareScenariosLimes` for information on how to call the function with the appropriate arguments.

Internally, `compareScenariosLimes()` calls `rmarkdown::render()` on certain Rmarkdown-files (Rmd-files). Rmd-files may contain R-code (e.g., for creating plots) as well as descriptive text (in this case, mostly section titles). The Rmd-files to be rendered are part of the limes-package. In the development state of the package, they can be found in the folder `inst/markdown/compareScenariosLimes/`. 
To be more precise, `rmarkdown::render()` is called on `cs2_main.Rmd` in this folder, which includes other Rmd-files - one for each section of the output document.
The loading and preprocessing of data happens in `cs2_main.Rmd`; the section Rmd-files mostly call plot functions.

Aside from HTML- and PDF-documents as output, `compareScenariosLimes()` also allows to obtain a copy of the Rmd-files needed to create these outputs (by setting the argument `outputFormat = "Rmd"`). Rendering the resulting Rmd-files to PDF or HTML yields the same documents as calls to `compareScenariosLimes()` with `outputFormat = "PDF"` or `"HTML"`. The advantage of being able to access these Rmd-files is the possibility to change the plotting code without changing the code of the `limes`-package. Moreover, Rmd-files can be used interactively in RStudio.

# Local Use and mif-Files

`compareScenariosLimes()` with default settings can be called automatically after a bundle of runs on the cluster has finished or manually via `Rscript output.R`. It can also be called directly as a function of the R-package `limes`, e.g.,

```{r}
library(limes)
compareScenariosLimes(
  mifScen = c("path/to/Base.mif", "path/to/NDC.mif"),
  outputFile = "compareScenariosLimesExample")
```

To do this locally on our computer, we need some mif files to compare. One possibility is to adapt and execute the following script to download the newest *modeltest* files.

```{r eval=FALSE}
library(tidyverse)
library(ssh)
# TODO: Adapt the following three lines!
keyfile <- "path/to/ssh/private/key/.ssh/id_rsa"
username <- "username"
out_path <- "path/to/where/mifs/will/be/downloaded/"
# path on the cluster where to search for mifs:
path <- "/p/projects/remind/modeltests/output/" 
# Connect to the cluster and get the file names of the newest modeltest mifs.
session <- ssh_connect(paste0(username, "@cluster.pik-potsdam.de"), keyfile=keyfile)
out <- ssh_exec_internal(session, paste("ls", path))
out$stdout %>%
  rawToChar() %>%
  str_split("\n", simplify = TRUE) %>%
  str_subset("^[a-zA-Z0-9-]+_\\d{4}-\\d{2}-\\d{2}_\\d{2}\\.\\d{2}\\.\\d{2}$") ->
  filenames
filenames %>%
  str_split("_", simplify = TRUE) %>%
  matrix(ncol=3, dimnames=list(NULL, c("name", "date", "time"))) %>%
  as_tibble() %>%
  bind_cols(filename = filenames) %>%
  mutate(date = as.Date(date)) %>%
  filter(date >= max(date)-2) ->
  new_file_data
new_files <- paste0(path, new_file_data$filename, "/LIMES_generic_", new_file_data$name, ".mif")
walk(new_files, scp_download, session = session, to = out_path) # download scenario mifs
hist_path <- paste0(path, new_file_data$filename[1], "/historical.mif") 
scp_download(session = session, hist_path, to = out_path) # download historical.mif
ssh_disconnect(session)
```


# Interactive Use of the Rmd-files

In this section, we assume that we do not want to change the package `limes`. But we may want to make changes to the compareScenarios output locally.

We call `limes::compareScenariosLimes()` with `outputFormat = "Rmd"` to obtain a (modified) copy of the Rmd-files.

```{r eval=FALSE}
library(limes)
compareScenariosLimes(
  mifScen = c("path/to/scen1.mif", "path/to/scen2.mif", "path/to/scen3.mif"), # TODO.
  outputDir = "path/to/where/rmds/should/be/copied/to", # TODO.
  outputFormat = "Rmd",
  # Add current time as output name to not overwrite other things:
  outputFile = format(Sys.time(), "CompScen2-%Y%m%d-%H%M%S")) 
```

The code in these Rmd-files is structured in chunks and each chunk can be run separately by clicking on its play-button. We first open `cs2_main.Rmd`. At the end of the file, there is a chunk
````
```{r prepare mark}
# CLICK "RUN ALL CHUNKS ABOVE" HERE TO PREPARE THE ENVIRONMENT
```
```` 
Clicking on the *triangle above bar*-button on the right executes all chunks above and by that loads and prepares the data for plotting. After that one can open any section-Rmd-file and execute a chunk to create the respective plots. The plots should appear in RStudio inside the Rmd-file below the chunk.

# Workflow for Adding a new Plot to compareScenariosLimes

1. If not done already, fork <https://github.com/pik-piam/limes>.
2. In your personal github limes repository, create a new branch.
3. Clone the repository and switch to the newly created branch to get your local copy.
4. In your local copy, open `limes.Rproj` in RStudio.
5. Press `CTRL+SHIFT+L` to call `devtools::load_all(".")`, which loads `limes` from your local copy.
6. Assuming you have some mifs available for testing, call

```{r eval=FALSE}
compareScenariosLimes(
  mifScen = c("path/to/scen1.mif", "path/to/scen2.mif"), # TODO.
  outputFile = format(Sys.time(), "CompScen2-%Y%m%d-%H%M%S"),
  sections = NULL,
  envir = globalenv())
```

This executes the code in `cs2_main.Rmd` without creating any plots (a mostly empty PDF-file will still be created). This will take some seconds, but it should not take minutes. Type `data` into the RStudio console and hit enter. This should show you the first rows of a table that contains all the loaded data (`ènvir = globalenv()` leads `render` to write all the variables to the environment available in the interactive session (the global environment)).

7. Open the section file in which you want to add the plot. The files are in the folder `inst/markdown/compareScenariosLimes/` and have names of the form `c2_NN_XXXX.Rmd`, where `NN` is a two digit number and `XXXX` resembles the section title.
8. You should be able to execute any chunk in the section Rmd by clicking the play button in its top right corner.
9. Insert a new chunk by copying an old one or by pressing `CTRL+ALT+I`.
10. Add a new plot inside the chunk, e.g., `showLinePlots(data, "Your|Variable|Name")`. See [Plot Functions] below.
11. Run your newly created chunk to see the plot.
12. Run the following code to see your new plot in a PDF.

```{r eval=FALSE}
compareScenariosLimes(
  mifScen = c("path/to/scen1.mif", "path/to/scen2.mif"), # TODO.
  outputFile = format(Sys.time(), "CompScen2-%Y%m%d-%H%M%S"),
  outputFormat = "PDF", # PDF is the default, so you can leave that line out. Or change it to html to get a html-version.
  sections = <NN>) # TODO: Replace <NN> by the number of the section you changed (or by "all" to build the whole compare scenarios file, which will take a while)
```

13. Save the file. Commit your changes. In the commit message, you may use the shorthand `cs2` to indicate that you made a change to compareScenariosLimes, e.g., `cs2: added plot of My|Variable`.
14. Make sure you fetched and pulled all recent changes to limes. Then, call `lucode2::buildLibrary()`. Commit again with the commit message `` `lucode2::buildLibrary()` ``. Push your commits to github.
15. Create a pull request.

# cs2_main.Rmd

This file loads the data from the mif-files, preprocesses the data, and includes the section-Rmd-files at the very end (and optionally further Rmd-files provided by the user, see YAML-parameter `userSectionPath`).

## YAML-header

The file `cs2_main.Rmd` starts with a YAML header marked by `---`. This header declares some basic information of the report, like its title and the output format. Furthermore, it contains a list `params`, which parameterizes the report. Among others, such parameters are the paths to the mif-files and certain properties that are shared for all plots in the report. Each such parameter can be changed by a respective argument in the call of `compareScenariosLimes()`, see section *YAML Parameters* in the documentation of the function.

## Loading

The mif-files are loaded using `quitte::read.quitte()`. This function names the global region `"World"` (not `"GLO"` as `magclass::read.report()`). For each file, we obtain one data-frame with columns `model, scenario, region, variable, unit, period, value`. During preprocessing the data-frames are concatenated into one quitte object named `data`. In the sections, `data` is used to provide the data necessary for the plots.

If provided, also config files are loaded.

## Preprocessing

* Scenarios are renamed if the user specifies new names (using a named vector for the argument `mifScen` of `compareScenariosLimes()` or by setting the parameter `mifScenNames` in the Rmd-files) or if scenario names are duplicate.
* The columns `period` (years) and `region` are filtered according to the parameters `yearsScen`, `yearsHist`, and `reg`.

## Global Variables 

Global variables are created in `cs2_main.Rmd` and are intended to be used in the plot functions of the section-Rmd-files.

The quitte-object (data frame) `data` with columns `model, scenario, region, variable, unit, period, value, varplus` provides all data that may be plotted.
It has mostly factor-columns. Thus, a vector of all available regions can be obtained by `levels(data$region)`.

Some arguments of the `mip::show...()` plot functions use the functionality provided by the base-R functions `options()` and `getOption()`. In `cs2_main.Rmd` there are calls to `options()`, setting values of the global options `mip.mainReg`, `mip.yearsBarPlot`, and `mip.histRefModel`. The plotting functions may have an argument, e.g., `mainReg` with default value `getOption("mip.mainReg")`. Thus, this argument does not have to be stated for each call of the plot function (assuming the respective call of `options()` sets it to the right value).

# Section-Rmd-files

The section-Rmd-files follow the naming pattern `cs2_NN_XXXX.Rmd`, where `NN` is replaced by a two digit number and `XXXX` is replaced by a short name of the section. If the YAML-parameter `sections` is set to `"all"`, the default, all sections of this naming pattern are included in `cs2_main.Rmd`. Alternatively, `sections` can be set to a vector of individual sections in the form of `"NN_XXXX"` to only render these sections.

The section-Rmd-files consist of section and sub-section titles, marked by `#`, `##`, `###`, ..., and R-code chunks which create plots, usually by calling one of the `show...()`-functions in the `mip` package.

## userSectionPath

The user can provide one or more additional Rmd-files that are appended after the sections provided by the package. Setting `sections` to `NULL` and `userSectionPath` to a character-vector of paths to Rmd-files creates a fully user-defined report.

## Special Sections

There is a first section (section number 0) and a last section in the files `cs2_00_info.Rmd` and `cs2_99_further_info.Rmd`, respectively. 

### 0 Info

This section contains a file reference table with scenario names and the paths to the respective mif files. It also show if a scenario is renamed. Renaming can be forced by the user by using a named character vector as argument `mifScen` where the names indicate the new scenario name, e.g.,

```{r}
compareScenariosLimes(
  mifScen = c(newName1 = "path/to/scen1.mif", newName2 = "path/to/scen2.mif"),
  ...)
```

Or, if two scenarios have the same name, they are renamed automatically.

Furthermore, this section displays a description of each scenario. This requires the YAML-parameter `cfgScen` to be set (see also [99 Further Info]). The description can be set in the `secenario_config*.csv`-file via a column `description`.


### 99 Further Info

If `compareScenariosLimes()` is provided with paths to the `config.Rdata` files of the scenarios via the YAML-parameter `cfgScen` and the path to `default.cfg` via `cfgDefault` an analysis of the configurations of the scenarios is shown.

```{r}
compareScenariosLimes(
  mifScen = c("path/to/scen1.mif", "path/to/scen2.mif"),
  cfgScen = c("path/to/scen1/config.yml", "path/to/scen2/config.yml"),
  cfgDefault = "path/to/default.cfg",
  ...)
```


# Plot Functions

The section Rmd-files mostly consists of simple calls of a `show...()`-function from the the [mip](https://github.com/pik-piam/mip) package. These functions are designed to make the creation of re-occurring kinds of plots in comapreSceanrios2 simple. Alternatively, one can build custom plots using `ggplot2`.

The `mip`-package contains following default plot functions for comapareScenarios2: 

* `showAreaAndBarPlots()`
* `showAreaAndBarPlotsPlus()`
* `showLinePlots()`
* `showLinePlotsWithTarget()`
* `showMultiLinePlots`
* `showMultiLinePlotsByVariable()`
* `showRegiLinePlots()`

All but `showLinePlotsWithTarget()` can use the global option `mip.mainReg`.
`showAreaAndBarPlots()` and `showAreaAndBarPlotsPlus()` can use `mip.yearsBarPlot`.
`showMultiLinePlotsByVariable()` uses `mip.histRefModel`.

See the respective function documentation for more information and example plots.
